Supplementary Information

### 2021-05-19 doc/manuscript/ForestGEO_treerings.Rmd
intro/ table 1 cleanup

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
Specifically in the case of drought, soil moisture (shaped by precipitation and evapotranspiration) and atmospheric vapor pressure deficit (VPD; driven by temperature) affect plant growth differently [@novick_increasing_2016; @humphrey_soil_2021], and we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].

Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @couvreur_water_2018; @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
Despite the fundamental importance of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

>>>>

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, commonly over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].

Tree diameter, DBH, scales predictably with numerous traits affecting tree physiology and growth [e.g., height, crown size and position, root mass, hydraulic architecture, @couvreur_water_2018; @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
Yet, there remain inconsistencies as to what is considered a typical growth pattern. 
Tree ring records commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas forest census data reveals a typical pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
Building upon observed ontogenetic patterns in RW, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
